---
title: "COVID-19 Analysis"
author: "Łukasz Pięta"
date: "11/11/2020"
output: 
  html_document: 
    keep_md: yes
    df_print: kable
    code_folding: "hide"
    toc: true
    toc_float: true
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


```{r libraries, include=FALSE}
library(readxl)
library(summarytools)
library(gtsummary)
library(tidyverse)
library(dplyr)
library(knitr)
library(DT)
library(ggplot2)
library(reshape2)
library(gganimate)
#library("gg3D")
library( rgl )
library(magick)
library(tibble)
```

## Read data
```{r read_data}
df <- read_excel("wuhan_blood_sample_data_Jan_Feb_2020.xlsx")
```

## Clean dataset
```{r, clean_dataset}

patients_df <- df %>% group_by(`Admission time`, `Discharge time`, gender, age, outcome) %>%
summarise(PATIENT_ID = sum(PATIENT_ID, na.rm = TRUE), `Total blood tests` = n()) 

patients_df <- patients_df %>%
    mutate(`Days in hospital` = ceiling(difftime(`Discharge time`, `Admission time`, units = "days")))


df <- full_join(patients_df %>% ungroup() %>% select(`Admission time`, PATIENT_ID), df %>% select(-PATIENT_ID), by="Admission time")

df$gender<-ifelse(df$gender==1, 'Male', 'Female') 
df <- df %>% mutate(gender = as.factor(gender))

patients_df$gender<-ifelse(patients_df$gender==1, 'Male', 'Female') 
patients_df <- patients_df %>% mutate(gender = as.factor(gender))

df$outcome<-ifelse(df$outcome==1, 'Death', 'Survival')
df <- df %>% mutate(outcome = as.factor(outcome))

patients_df$outcome<-ifelse(patients_df$outcome==1, 'Death', 'Survival')
patients_df <- patients_df %>% mutate(outcome = as.factor(outcome))
```

## Patients summary
```{r, data_transformed, warning=FALSE}

patients_summary <- patients_df %>% ungroup() %>% select(-c(age, PATIENT_ID))

tbl_summary(
    patients_summary,
    by = outcome,
    label = gender ~ "Gender",
) %>%
    add_n() %>%
    modify_header(label = "") %>%
    add_overall() %>%
    bold_labels() 

 ggplot(patients_df, aes(x=age,fill=gender)) + geom_histogram(binwidth = 1) + facet_grid(. ~ gender) + scale_x_continuous(name="Age", limits=c(min(df$age), max(df$age)), breaks = seq(0, 100, by=10)) + scale_y_continuous(name = "Number of patients", limits = c(0,10), breaks = seq(0,10, by=1)) +
    theme_minimal()
```

### Patients grouped by outcome, age and gender
```{r, gender_distribution, warning=FALSE}
 ggplot(patients_df, aes(x=age,fill=outcome)) + geom_histogram(binwidth = 1) + facet_grid(outcome ~ gender) + scale_x_continuous(name="Age", limits=c(min(df$age), max(df$age)), breaks = seq(0, 100, by=10)) + scale_y_continuous(name = "Number of patients", limits = c(0,9), breaks = seq(0,9, by=1)) +
    theme_minimal()
```

### Patients grouped by outcome and hospitalization duration
```{r, hospitalization_duration, warning=FALSE}

ggplot(patients_df, aes(x=`Days in hospital`, fill=outcome)) + geom_histogram(binwidth = 1) + facet_grid(. ~ outcome)  + ylab("Number of patients") +
    theme_minimal()

```

### Patients grouped by outcome and total blood tests taken
```{r, total_blood_tests, warning=FALSE}

ggplot(patients_df, aes(x=`Total blood tests`, fill=outcome)) + geom_histogram(binwidth = 1) + facet_grid(. ~ outcome)  + ylab("Number of patients") +
  scale_x_continuous(name="Total blood tests", limits=c(0, 60), breaks = seq(0, 60, by=10)) +
   scale_y_continuous(name = "Number of patients", limits = c(0,20), breaks = seq(0,20, by=2)) +
    theme_minimal()

```

## Biomarkers
```{r original_data_summary,message=FALSE,warning=FALSE}

no_dates_df <- df %>% select(-c('Admission time', 'Discharge time', 'RE_DATE', 'PATIENT_ID', 'age', 'gender' ))
tbl_summary(
    no_dates_df,
    by = outcome,
    missing = "no"
            ) %>%
    add_n() %>%
    modify_header(label = "**Biomarker**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Final patient outcome related to the test**") %>%
    bold_labels() 

```

### Finding correlation

```{r, find_cor}
cor_df <-df

cor_df$outcome<-ifelse(cor_df$outcome=='Death', 1, 0)
cor_df$gender<-ifelse(cor_df$gender=='Male', 1, 0)

cor_df <- cor_df %>% rename(isMale=gender)
cor_df <- cor_df %>% rename(Death=outcome)

#cor_df %>% select (-c(`Admission time`, PATIENT_ID, RE_DATE, `Discharge time` ))

cor_df <- cor_df %>% select (-c(`Admission time`, PATIENT_ID, RE_DATE, `Discharge time` ))

corrMatrix <- cor(cor_df[sapply(cor_df, is.numeric)], use='pairwise.complete.obs')

correlation_df <- as.data.frame(corrMatrix)

#correlation_df[3,]

correlation_df %>% rownames_to_column('variable') %>% filter(variable != 'Death') %>% select(variable, Death) %>% mutate(Death = abs(Death)) %>%
  arrange(desc(Death)) %>%
  rename(`Outcome correlation` = Death) %>%
  head(10) 

#highlyCorrelated <- findCorrelation(corrMatrix, cutoff=0.5)

```

Taking only these patients that have all of the following biomarkers tested: Lactate dehydrogenase, Lymphocyte count, High sensitivity C-reactive protein.

```{r, correlation ,webgl=TRUE,results='hide'}

new_df  <- df

new_df$outcome<-ifelse(df$outcome=='Death', 1, 0)

new_df <- new_df %>% select(PATIENT_ID, `Lactate dehydrogenase`, `lymphocyte count`, `High sensitivity C-reactive protein`, outcome) %>%
    group_by(PATIENT_ID) %>%
    summarise(LDH_mean = mean(`Lactate dehydrogenase`, na.rm = TRUE), Lymphocyte_mean = mean(`lymphocyte count`, na.rm = TRUE), hs_CRP_mean = mean(`High sensitivity C-reactive protein`, na.rm = TRUE), death = sum(outcome) > 0)

new_df <- new_df %>% filter(!is.nan(LDH_mean) & !is.nan(Lymphocyte_mean) & !is.nan(hs_CRP_mean))

new_df$death<-ifelse(new_df$death==TRUE, 'Death', 'Survival')
new_df <- new_df %>% mutate(death = as.factor(death)) %>% rename(outcome = death)


mycolors <- c('royalblue1', 'darkcyan')
new_df$color <- mycolors[ as.numeric(new_df$outcome) ]

par(mar=c(0,0,0,0))
plot3d( 
    x=new_df$LDH_mean, y=new_df$Lymphocyte_mean, z=new_df$hs_CRP_mean, 
    col = new_df$color, 
    type = 's', 
    radius = 10,
    legend=TRUE,
    xlab="LDH_mean", ylab="Lymphocyte_mean", zlab="hs_CRP_mean")
legend3d("topright", legend = c('Death', 'Survival'), pch = 10, col = mycolors, cex=0.8, inset=c(0.02))

writeWebGL( filename="3dscatter.html" ,  width=600, height=600)

#p <- ggplot((df %>% filter(!is.na())), aes(RE_DATE, `Lactate dehydrogenase`, colour = outcome)) +
#    geom_point(alpha = 0.7) +
#   labs(title = 'PATIENT_ID: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
#    transition_time(PATIENT_ID) +
#    ease_aes('linear')

#animate(p, 375, 1)
```


Patients summary then:

```{r, patients_with_markers} 

tbl_summary(
    new_df %>% rename(`Lactate dehydrogenase` = LDH_mean, `Lymphocyte count` = Lymphocyte_mean, `High sensitivity C-reactive protein`= hs_CRP_mean) %>% select(-PATIENT_ID, -color),
    by = outcome
) %>%
    add_n() %>%
    modify_header(label = "**Mean value of**") %>%
    add_overall() %>%
    bold_labels() 

```



Correlation between lactate dehydrogenase, lymphocyte count and high sensitivity C-reactive protein:
```{r, show_correlation}

htmltools::includeHTML("./3dscatter.html")

#df %>% filter(!is.na(`Lactate dehydrogenase`)) %>% group_by(PATIENT_ID) %>% summarise(LDH_mean = mean(`Lactate dehydrogenase`), survived = sum(as.numeric(outcome)) > 0)
```

