---
title: "COVID-19 Analysis"
author: "Łukasz Pięta"
date: "11/11/2020"
output: 
  html_document: 
    keep_md: yes
    df_print: kable
    code_folding: "hide"
    toc: true
    toc_float: true
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


```{r libraries, include=FALSE}
library(readxl)
library(summarytools)
library(gtsummary)
library(tidyverse)
library(dplyr)
library(knitr)
library(DT)
library(ggplot2)
library(reshape2)
library(gganimate)
library(hrbrthemes)
library( rgl )
library(magick)
library(tibble)
library(caret)
```

## Read data
```{r read_data}
df <- read_excel("wuhan_blood_sample_data_Jan_Feb_2020.xlsx")
```

## Clean dataset
```{r, clean_dataset}

patients_df <- df %>% group_by(`Admission time`, `Discharge time`, gender, age, outcome) %>%
summarise(PATIENT_ID = sum(PATIENT_ID, na.rm = TRUE), `Total records` = n()) 

patients_df <- patients_df %>%
    mutate(`Days in hospital` = ceiling(difftime(`Discharge time`, `Admission time`, units = "days")))

df <- full_join(patients_df %>% ungroup() %>% select(`Admission time`, PATIENT_ID), df %>% select(-PATIENT_ID), by="Admission time")

df$gender<-ifelse(df$gender==1, 'Male', 'Female') 
df <- df %>% mutate(gender = as.factor(gender))

patients_df$gender<-ifelse(patients_df$gender==1, 'Male', 'Female') 
patients_df <- patients_df %>% mutate(gender = as.factor(gender))

df$outcome<-ifelse(df$outcome==1, 'Death', 'Survival')
df <- df %>% mutate(outcome = as.factor(outcome))

patients_df$outcome<-ifelse(patients_df$outcome==1, 'Death', 'Survival')
patients_df <- patients_df %>% mutate(outcome = as.factor(outcome))

patients_df <- df %>% group_by(PATIENT_ID, outcome, `Admission time`, `Discharge time`, gender, age) %>% summarise(missing_test=is.na(RE_DATE)) %>%
  group_by(PATIENT_ID, outcome, `Admission time`, `Discharge time`, gender, age) %>%
  summarise(test_provided = ifelse(missing_test, 0, 1)) %>%
  group_by(PATIENT_ID, outcome, `Admission time`, `Discharge time`, gender, age) %>%
  summarise(`Total blood tests`=sum(test_provided)) %>%
  mutate(`Days in hospital` = ceiling(difftime(`Discharge time`, `Admission time`, units = "days")))

last_test_df <- df %>% 
  group_by(PATIENT_ID, outcome, gender, age) %>% 
  fill_(names(df)) %>% 
  fill_(names(df), "up") %>% 
  summarise_at(vars(`Hypersensitive cardiac troponinI`:creatinine), function(x) last(x,order_by = is.na(x)))


cleaned_df <- last_test_df %>% ungroup() %>% select(PATIENT_ID, outcome, age, gender, hemoglobin, `eosinophils(%)`, `Alkaline phosphatase`, albumin, `basophil(%)`, `Total bilirubin`, `Platelet count`, `monocytes(%)`, `neutrophils(%)`, `total protein`, `mean corpuscular volume`, hematocrit, `White blood cell count`, `mean corpuscular hemoglobin concentration`, Urea, `lymphocyte count`, `Red blood cell count`, `Eosinophil count`, `neutrophils count`, `Direct bilirubin`, `(%)lymphocyte`, `Total cholesterol`, `aspartate aminotransferase`, `Uric acid`, `HCO3-`, `Lactate dehydrogenase`, `monocytes count`, globulin, `γ-glutamyl transpeptidase`, `basophil count(#)`, `mean corpuscular hemoglobin`, `glutamic-pyruvic transaminase`, eGFR, creatinine) %>% filter_all(function(x) !is.na(x)) 
```

## Patients summary
```{r, data_transformed, warning=FALSE}

patients_summary <- patients_df %>% ungroup() %>% select(-c(age, PATIENT_ID))

tbl_summary(
    patients_summary,
    by = outcome,
    label = gender ~ "Gender",
) %>%
    add_n() %>%
    modify_header(label = "") %>%
    add_overall() %>%
    bold_labels() 

 ggplot(patients_df, aes(x=age,fill=gender)) + geom_histogram(binwidth = 1) + facet_grid(. ~ gender) + scale_x_continuous(name="Age", limits=c(min(df$age), max(df$age)), breaks = seq(0, 100, by=10)) + scale_y_continuous(name = "Number of patients", limits = c(0,10), breaks = seq(0,10, by=1)) +
    theme_minimal()
```

### Patients grouped by outcome, age and gender
```{r, gender_distribution, warning=FALSE}
 ggplot(patients_df, aes(x=age,fill=outcome)) + geom_histogram(binwidth = 1) + facet_grid(outcome ~ gender) + scale_x_continuous(name="Age", limits=c(min(df$age), max(df$age)), breaks = seq(0, 100, by=10)) + scale_y_continuous(name = "Number of patients", limits = c(0,9), breaks = seq(0,9, by=1)) +
    theme_minimal()
```

### Patients grouped by outcome and hospitalization duration
```{r, hospitalization_duration, warning=FALSE}

ggplot(patients_df, aes(x=`Days in hospital`, fill=outcome)) + geom_histogram(binwidth = 1) + facet_grid(. ~ outcome)  + ylab("Number of patients") +
    theme_minimal()

```

### Patients grouped by outcome and total blood tests taken
```{r, total_blood_tests, warning=FALSE}

ggplot(patients_df, aes(x=`Total blood tests`, fill=outcome)) + geom_histogram(binwidth = 1) + facet_grid(. ~ outcome)  + ylab("Number of patients") +
  scale_x_continuous(name="Total blood tests", limits=c(0, 60), breaks = seq(0, 60, by=10)) +
   scale_y_continuous(name = "Number of patients", limits = c(0,20), breaks = seq(0,20, by=2)) +
    theme_minimal()

```

## Removed data

```{r, removed_data}
patients_df %>% filter(`Total blood tests`==0) 
```

## Biomarkers
```{r original_data_summary,message=FALSE,warning=FALSE}

no_dates_df <- df %>% select(-c('Admission time', 'Discharge time', 'RE_DATE', 'PATIENT_ID', 'age', 'gender' ))
tbl_summary(
    no_dates_df,
    by = outcome,
    missing = "no"
            ) %>%
    add_n() %>%
    modify_header(label = "**Biomarker**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Final patient outcome related to the test**") %>%
    bold_labels() 

```

### Finding correlation

Finding correlation between outcome and other variables. Ignoring `Admission time`, `PATIENT_ID`, `RE_DATE` and `Discharge time`, and dropping `outcome`, which correlation with itself is equal to `1.0`.

```{r, find_cor}
cor_df <-df

cor_df$outcome<-ifelse(cor_df$outcome=='Death', 1, 0)
cor_df$gender<-ifelse(cor_df$gender=='Male', 1, 0)

cor_df <- cor_df %>% rename(isMale=gender)
cor_df <- cor_df %>% rename(Death=outcome)

cor_df <- cor_df %>% select (-c(`Admission time`, PATIENT_ID, RE_DATE, `Discharge time` ))

corrMatrix <- cor(cor_df[sapply(cor_df, is.numeric)], use='pairwise.complete.obs')

correlation_df <- as.data.frame(corrMatrix)

correlation_df %>% rownames_to_column('variable') %>% filter(variable != 'Death') %>% select(variable, Death) %>% mutate(Death = abs(Death)) %>%
  arrange(desc(Death)) %>%
  rename(`Outcome correlation` = Death) %>%
  head(10) 

```

### Use of correlation

In following section, we will use 3 biomarkers that are most correlated to outcome and visualize theirs mean values among each patient at 3D graph. So we take only these patients who had all of these 3 biomarkers tested at least once. If somebody was tested more than once, then mean value of all tests for each biomarker is calculated.

```{r, use_correlation_mean_value ,webgl=TRUE,results='hide'}

#TODO: Take for correlation cleaned_df and check if it's consistent with predictive model

new_df  <- df

new_df <- new_df %>% select(PATIENT_ID, `neutrophils(%)`, `(%)lymphocyte`, albumin, outcome) %>%
    group_by(PATIENT_ID, outcome) %>%
    summarise(Neutrophils_mean = mean(`neutrophils(%)`, na.rm = TRUE), Lymphocyte_mean = mean(`(%)lymphocyte`, na.rm = TRUE), albumin_mean = mean(`albumin`, na.rm = TRUE))

new_df <- new_df %>% filter(!is.nan(Neutrophils_mean) & !is.nan(Lymphocyte_mean) & !is.nan(albumin_mean))


mycolors <- c('royalblue1', 'darkcyan')
new_df$color <- mycolors[ as.numeric(new_df$outcome) ]

par(mar=c(0,0,0,0))
plot3d( 
    x=new_df$Neutrophils_mean, y=new_df$Lymphocyte_mean, z=new_df$albumin_mean, 
    col = new_df$color, 
    type = 's', 
    radius = 1,
    legend=TRUE,
    xlab="Neutrophils(%)", ylab="Lymphocyte(%)", zlab="Albumin")
legend3d("topright", legend = c('Death', 'Survival'), pch = 10, col = mycolors, cex=0.8, inset=c(0.02))

writeWebGL( filename="3d_correlation_mean.html" ,  width=600, height=600)
```

Data after cleaning and transformation:

```{r, correlation_summary_mean} 

tbl_summary(
    new_df %>% ungroup() %>%rename(`Neutrophils(%)` = Neutrophils_mean, `Lymphocyte(%)` = Lymphocyte_mean, `Albumin`= albumin_mean) %>% select(-PATIENT_ID, -color),
    by = outcome
) %>%
    add_n() %>%
    modify_header(label = "**Mean value of**") %>%
    add_overall() %>%
    bold_labels() 

```

```{r, show_correlation_mean}
htmltools::includeHTML("./3d_correlation_mean.html")
```

## Common biomarkers' change in time

```{r, most_common_biomarkers} 
biomarker_popularity <- df %>% group_by(PATIENT_ID, outcome) %>% summarise_at(vars(`Hypersensitive cardiac troponinI`:creatinine), function(x) sum(!is.na(x)))
biomarker_popularity$sum_of_single_test <- rowSums(biomarker_popularity %>% ungroup() %>% select(-PATIENT_ID, -outcome))

most_tested_death <- biomarker_popularity %>% select(sum_of_single_test, outcome, `neutrophils(%)`,  `(%)lymphocyte`, albumin) %>% filter(outcome=='Death') %>% arrange(desc(sum_of_single_test)) %>% head(10)
most_tested_survival <- biomarker_popularity %>% select(sum_of_single_test, outcome, `neutrophils(%)`,  `(%)lymphocyte`, albumin) %>% filter(outcome=='Survival') %>% arrange(desc(sum_of_single_test)) %>% head(10)

foo <- rbind(most_tested_death, most_tested_survival)

patients_tests <- merge(foo %>% ungroup() %>% select(PATIENT_ID), df %>% ungroup() %>% select(PATIENT_ID, RE_DATE,`neutrophils(%)`,  `(%)lymphocyte`, albumin, outcome),  by="PATIENT_ID") %>% filter(!is.na(`neutrophils(%)`) |  !is.na(`(%)lymphocyte`) | !is.na(albumin))

neutrophils_seq <- patients_tests %>% filter(!is.na(`neutrophils(%)`)) 

neutrophils_seq %>% mutate(PATIENT_ID=as.factor(PATIENT_ID)) %>%
  ggplot( aes(x=RE_DATE, y=`neutrophils(%)`, group=PATIENT_ID, color=PATIENT_ID)) +
    geom_line() +
    geom_point() +
  facet_grid(rows=vars(outcome)) +
    ggtitle("Neutrophils (%) during patient hospitalization") +
    theme_ipsum() +
    ylab("Neutrophils (%)") +
    transition_reveal(RE_DATE)

#anim_save("neutrophils_seq.gif")

lymphocyte_seq <- patients_tests %>% filter(!is.na(`(%)lymphocyte`)) 

lymphocyte_seq %>% mutate(PATIENT_ID=as.factor(PATIENT_ID)) %>%
  ggplot( aes(x=RE_DATE, y=`(%)lymphocyte`, group=PATIENT_ID, color=PATIENT_ID)) +
    geom_line() +
    geom_point() +
  facet_grid(rows=vars(outcome)) +
    ggtitle("Lymphocyte (%) during patient hospitalization") +
    theme_ipsum() +
    ylab("Lymphocyte (%)") +
    transition_reveal(RE_DATE)

#anim_save("lymphocyte_seq.gif")

albumin_seq <- patients_tests %>% filter(!is.na(albumin)) 

albumin_seq %>% mutate(PATIENT_ID=as.factor(PATIENT_ID)) %>%
  ggplot( aes(x=RE_DATE, y=albumin, group=PATIENT_ID, color=PATIENT_ID)) +
    geom_line() +
    geom_point() +
  facet_grid(rows=vars(outcome)) +
    ggtitle("Albumin value during patient hospitalization") +
    theme_ipsum() +
    ylab("Albumin") +
    transition_reveal(RE_DATE)

#anim_save("albumin_seq.gif")

```

## Prediction model

In order to build an appropriate model, there is a need for the data cleaning. There shouldn't be any `NA` in the data set. Moreover, every considered patient should have all of considered biomarkers tested at least once. For building the model purpose, the last existing test of each biomarker for each patient was taken. As it was shown before, many biomarkers are unsuitable because of containing too many `NA` in theirs columns. Which means that they should be removed from the data set. Below is the list of remaining columns:

<ul>
<b>
  <li>outcome</li>
  <li>age</li>
  <li>gender</li>
  <li>hemoglobin</li>
  <li>eosinophils(%)</li>
  <li>Alkaline phosphatas</li>
  <li>albumin</li>
  <li>basophil(%)</li>
  <li>Total bilirubin</li>
  <li>Platelet count</li>
  <li>monocytes(%)</li>
  <li>neutrophils(%)</li>
  <li>total protein</li>
  <li>mean corpuscular volume</li>
  <li>hematocrit</li>
  <li>White blood cell count</li>
  <li>mean corpuscular hemoglobin concentration</li>
  <li>Urea</li>
  <li>lymphocyte count</li>
  <li>Red blood cell count</li>
  <li>Eosinophil count</li>
  <li>neutrophils count</li>
  <li>Direct bilirubin</li>
  <li>(%)lymphocyte</li>
  <li>Total cholesterol</li>
  <li>aspartate aminotransferase</li>
  <li>Uric acid</li>
  <li>HCO3-</li>
  <li>Lactate dehydrogenase</li>
  <li>monocytes count</li>
  <li>globulin</li>
  <li>γ-glutamyl transpeptidase</li>
  <li>basophil count(#)</li>
  <li>mean corpuscular hemoglobin</li>
  <li>glutamic-pyruvic transaminase</li>
  <li>eGFR</li>
  <li>creatinine</li>
  </b>
</ul>  

With condition of existing at least one of listed above biomarkers tests, 19 patients should be ignored in building model process.

```{r,prediction_model} 
cleaned_df <- cleaned_df %>% select(-PATIENT_ID)
set.seed(23)
inTraining <- 
    createDataPartition(
        y = cleaned_df$outcome,
        p = .70,
        list = FALSE)


training <- cleaned_df[ inTraining,]
testing  <- cleaned_df[-inTraining,]

rfGrid <- expand.grid(mtry = 10:30)
ctrl <- trainControl(
    method = "repeatedcv",
    classProbs = TRUE,
    number = 2,
    repeats = 5)

set.seed(23)
fit <- train(outcome ~ .,
             data = training,
             method = "rf",
             metric = "ROC",
             preProc = c("center", "scale"),
             trControl = ctrl,
             tuneGrid = rfGrid,
             ntree = 30)
rfClasses <- predict(fit, newdata = testing)
confusionMatrix(data = rfClasses, testing$outcome)
```


### Imporance of attributes

```{r,importance_attributes} 
varImp(fit)
```